---
title: "Lab6"
format: html
editor: visual
---

#Question 1: Download data
##Set Up

```{r}
library(tidyverse)
library(tidymodels)
library(powerjoin)
library(glue)
library(vip)
library(baguette)
library(patchwork)
```

```{r}
root  <- 'https://gdex.ucar.edu/dataset/camels/file'
download.file('https://gdex.ucar.edu/dataset/camels/file/camels_attributes_v2.0.pdf', 
              'data/camels_attributes_v2.0.pdf', 
              mode = 'wb')

```
```{r}
types <- c("clim", "geol", "soil", "topo", "vege", "hydro")

# Where the files live online ...
remote_files  <- glue('{root}/camels_{types}.txt')
# where we want to download the data ...
local_files   <- glue('data/camels_{types}.txt')

walk2(remote_files, local_files, download.file, quiet = TRUE)

# Read and merge data
camels <- map(local_files, read_delim, show_col_types = FALSE)

# merge the data into a single data frame
camels <- reduce(camels, power_full_join, by = 'gauge_id')

```

```{r}
list.files("data")
```
###yay it worked!  all 6 data files + the PDF are here

#Question 2:Make 2 maps

```{r}
p1 <- ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = q_mean)) +
  scale_color_gradient(low = "pink", high = "dodgerblue") +
  ggthemes::theme_map()

p1
```

###It looks just like the map in the assignment!  
```{r}
# First map: Color points by aridity
p2 <- ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = aridity)) +
  scale_color_gradient(low = "pink", high = "navy") +
  labs(color = "Aridity Index") +
  ggthemes::theme_map()

# Second map: Color points by p_mean (mean annual precipitation)
p3 <- ggplot(data = camels, aes(x = gauge_lon, y = gauge_lat)) +
  borders("state", colour = "gray50") +
  geom_point(aes(color = p_mean)) +
  scale_color_gradient(low = "lightblue", high ="purple") +
  labs (color = "p_mean (mm)") +
  ggthemes::theme_map()


p2 + p3
```

#Question 3: Build a xgboost and neural network model 
##Model Preparation

```{r}
camels |> 
  select(aridity, p_mean, q_mean) |> 
  drop_na() |> 
  cor()
```
###Looks like the expected results from the assignments

```{r}
# Create a scatter plot of aridity vs rainfall
ggplot(camels, aes(x = aridity, y = p_mean)) +
  # Add points colored by mean flow
  geom_point(aes(color = q_mean)) +
  # Add a linear regression line
  geom_smooth(method = "lm", color = "red", linetype = 2) +
  # Apply the viridis color scale
  scale_color_viridis_c() +
  # Add a title, axis labels, and theme (w/ legend on the bottom)
  theme_linedraw() + 
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runnoff", 
       x = "Aridity", 
       y = "Rainfall",
       color = "Mean Flow")
```
###The relationship is seen here just like in the assignment.

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  scale_color_viridis_c() +
  # Apply log transformations to the x and y axes
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom") + 
  labs(title = "Aridity vs Rainfall vs Runnoff", 
       x = "Aridity", 
       y = "Rainfall",
       color = "Mean Flow")
```
###Log scaled made a more leniar relationship.

```{r}
ggplot(camels, aes(x = aridity, y = p_mean)) +
  geom_point(aes(color = q_mean)) +
  geom_smooth(method = "lm") +
  # Apply a log transformation to the color scale
  scale_color_viridis_c(trans = "log") +
  scale_x_log10() + 
  scale_y_log10() +
  theme_linedraw() +
  theme(legend.position = "bottom",
        # Expand the legend width ...
        legend.key.width = unit(2.5, "cm"),
        legend.key.height = unit(.5, "cm")) + 
  labs(title = "Aridity vs Rainfall vs Runnoff", 
       x = "Aridity", 
       y = "Rainfall",
       color = "Mean Flow") 
```
###Definitely more evenly spread just like expected.

```{r}
set.seed(123)
# Bad form to perform simple transformations on the outcome variable within a 
# recipe. So, we'll do it here.
camels <- camels |> 
  mutate(logQmean = log(q_mean))

# Generate the split
camels_split <- initial_split(camels, prop = 0.8)
camels_train <- training(camels_split)
camels_test  <- testing(camels_split)

camels_cv <- vfold_cv(camels_train, v = 10)
```

```{r}
# Create a recipe to preprocess the data
rec <-  recipe(logQmean ~ aridity + p_mean, data = camels_train) %>%
  # Log transform the predictor variables (aridity and p_mean)
  step_log(all_predictors()) %>%
  # Add an interaction term between aridity and p_mean
  step_interact(terms = ~ aridity:p_mean) |> 
  # Drop any rows with missing values in the pred
  step_naomit(all_predictors(), all_outcomes())
```

```{r}
# Prepare the data
baked_data <- prep(rec, camels_train) |> 
  bake(new_data = NULL)

# Interaction with lm
#  Base lm sets interaction terms with the * symbol
lm_base <- lm(logQmean ~ aridity * p_mean, data = baked_data)
summary(lm_base)
```

###Just as expected

```{r}
# Sanity Interaction term from recipe ... these should be equal!!
summary(lm(logQmean ~ aridity + p_mean + aridity_x_p_mean, data = baked_data))
```


###Just as expected again

```{r}
nrow(camels_test)
nrow(camels_train)

```
```{r}
broom::augment(lm_base, data = camels_test)
```
###I was worried there for a second, it looks like an error was expected

```{r}
camels_test$p2 = predict(lm_base, newdata = camels_test)

## Scales way off!
ggplot(camels_test, aes(x = p2, y = logQmean)) + 
  geom_point() + 
  # Linear fit line, no error bands
  geom_smooth(method = "lm", se = FALSE, size =1) +
  # 1:1 line
  geom_abline(color = "red", size = 1) + 
  labs(title = "Linear Model Using `predict()`",
       x = "Predicted Log Mean Flow",
       y = "Observed Log Mean Flow") + 
  theme_linedraw()
```
###This is just like the assignment!

```{r}
test_data <-  bake(prep(rec), new_data = camels_test)
test_data$lm_pred <- predict(lm_base, newdata = test_data)
```

```{r}
metrics(test_data, truth = logQmean, estimate = lm_pred)
```

```{r}
ggplot(test_data, aes(x = logQmean, y = lm_pred, colour = aridity)) +
  # Apply a gradient color scale
  scale_color_gradient2(low = "pink", mid = "purple", high = "navy") +
  geom_point() +
  geom_abline(linetype = 2) +
  theme_linedraw() + 
  labs(title = "Linear Model: Observed vs Predicted",
       x = "Observed Log Mean Flow",
       y = "Predicted Log Mean Flow",
       color = "Aridity")
```
###Looks the same just with changed colors since the other looked a bit off to me.
###This really was a lot of work for one single graph...

##Better approach: Workflow

```{r}
# Define model
lm_model <- linear_reg() %>%
  # define the engine
  set_engine("lm") %>%
  # define the mode
  set_mode("regression")

# Instantiate a workflow ...
lm_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec) %>%
  # Add the model
  add_model(lm_model) %>%
  # Fit the model to the training data
  fit(data = camels_train) 

# Extract the model coefficients from the workflow
summary(extract_fit_engine(lm_wf))$coefficients
```
```{r}
# From the base implementation
summary(lm_base)$coefficients
```
###Now workflow is embedded in the model!

##Make Predictions:

```{r}
#
lm_data <- augment(lm_wf, new_data = camels_test)
dim(lm_data)
```
```{r}
metrics(lm_data, truth = logQmean, estimate = .pred)
```

```{r}
ggplot(lm_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```
###Boom! an easy to make and more adaptble graph from my data!

```{r}
library(baguette)
rf_model <- rand_forest() %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

rf_wf <- workflow() %>%
  # Add the recipe
  add_recipe(rec) %>%
  # Add the model
  add_model(rf_model) %>%
  # Fit the model
  fit(data = camels_train) 
```

```{r}
rf_data <- augment(rf_wf, new_data = camels_test)
dim(rf_data)
```
```{r}
metrics(rf_data, truth = logQmean, estimate = .pred)
```

```{r}
ggplot(rf_data, aes(x = logQmean, y = .pred, colour = aridity)) +
  scale_color_viridis_c() +
  geom_point() +
  geom_abline() +
  theme_linedraw()
```
### Here is the proof, using the framwork for a completely new model!

##Workflowset approach

```{r}
wf <- workflow_set(list(rec), list(lm_model, rf_model)) %>%
  workflow_map('fit_resamples', resamples = camels_cv) 

autoplot(wf)
```
```{r}
rank_results(wf, rank_metric = "rsq", select_best = TRUE)
```
#Question 4a: Data Prep / Data Splitting
#Question 4b: Recipe 
#Question 4c: Define 3 models 
#Question 4d: workflow set 
#Question 4e: Evaluation 
#Question 4f: Extract and Evaluate 
